---
title: "Climate data and stations"
author: "Sanne Evers"
date: "4/28/2022"
output: pdf_document
params:
  gps_locations: "data/Draco_locations.csv"
  data_loggers_folder: "data/clima/"
  climate_dobrichovice: "data/climatic data Dobrichovice.xlsx"
  chelsa_data: "data/CHELSA_data.csv"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F)

suppressPackageStartupMessages(library(leaflet))
suppressPackageStartupMessages(library(tidyverse))
```

# CZ climate stations

```{r map, echo= FALSE, warning=FALSE}

df <- read.csv(here::here(params$gps_locations)) %>%
  rename(id = `Station.ID.Draco.locality.ID`,
         lat = Latitude,
         lon = Longitude) %>%
  mutate(type = if_else(id %in% c("Hk", "Cr", "Ks", "Ru"), "population", "station"))


pal <- colorFactor(c("navy", "red"), domain = c("station", "population"), ordered = T)


leaflet(df) %>%
  setView(lng = 14.2, lat = 49.95, zoom = 11) %>%
  addProviderTiles(providers$Esri.WorldStreetMap) %>% 
  addCircleMarkers(
    color = ~pal(type),
    fillOpacity = 1,
    stroke = FALSE,
    popup = ~htmltools::htmlEscape(Name)) %>%
  addLegend("bottomright",
            colors = c("navy", "red"),
            labels = c("station", "population"),
            values = ~id, opacity = 1, title = "Locations") %>%
  addScaleBar()

```


## Timeseries of, and correlations between, weather in climate stations

```{r 1, echo= FALSE, warning=FALSE}

clim <- readxl::read_xlsx(here::here(params$climate_dobrichovice)) %>%
  group_by(Year, Month) %>%
  summarise(Prec_Dobrichovice = sum(Prec_Dobrichovice),
            Prec_Lodenice = sum(Prec_Lodenice),
            Prec_Beroun = sum(Prec_Beroun),
            Prec_Liten = sum(Prec_Liten)) %>%
  ungroup(Year) %>%
  mutate(across(contains("Prec"), ~ scale(.x))) %>%
  pivot_longer(cols = contains("Prec"), names_to = "station", values_to = "sum_prcp") %>%
  ungroup()

ggplot(clim) +
  geom_line(aes(x = Month, y = sum_prcp, colour = station)) + 
  facet_wrap(vars(Year))

```

Above are the climate (anomalies) plotted across time for all 4 climate stations. As you can see there is a lot of overlap. However, a lot of times when there is no overlap, there's an "extreme" peak that would be very valuable for the climate analysis, if different localities also respond to different stations. 

Below we have the correlation matrix between the 4 different stations.

```{r 1.5, echo=FALSE, warning=FALSE}

pivot_wider(clim, names_from = station, values_from = sum_prcp) %>%
  dplyr::select(-c(Year, Month)) %>%
  cor(.)


```

# CHELSA climate
With the distance between the climate stations and localities it doesn't really make sense to use different stations for each of the populations. However, there's also enough distance between the populations that it might be worthwhile using different climate for each of the localities. Using the gps locations of the localities I retrieved climate for these locations from CHELSA.


```{r}
chelsa <- read.csv(here::here(params$chelsa_data))

print("precipitation")
chelsa %>% 
  dplyr::select(Name, month, year, pr) %>%
  filter(Name %in% c("Haknovec", "Císarská rokle", "Kodská stena", "Radotánské údolí")) %>%
  pivot_wider(names_from = Name, values_from = pr) %>% 
  dplyr::select(-c(year, month)) %>%
  filter(complete.cases(.)) %>%
  cor(.)

```

That really is very high! What about the other climate variables?

```{r}
print("temperature")
chelsa %>% 
  dplyr::select(Name, month, year, tas) %>%
  filter(Name %in% c("Haknovec", "Císarská rokle", "Kodská stena", "Radotánské údolí")) %>%
  pivot_wider(names_from = Name, values_from = tas) %>% 
  dplyr::select(-c(year, month)) %>%
  filter(complete.cases(.)) %>%
  cor(.)

# print("climate moisture index")
# chelsa %>% 
#   dplyr::select(Name, month, year, cmi) %>%
#   filter(Name %in% c("Haknovec", "Císarská rokle", "Kodská stena", "Radotánské údolí")) %>%
#   pivot_wider(names_from = Name, values_from = cmi) %>% 
#   dplyr::select(-c(year, month)) %>%
#   filter(complete.cases(.)) %>%
#   cor(.)

print("Potential evapotranspiration")
chelsa %>% 
  dplyr::select(Name, month, year, pet) %>%
  filter(Name %in% c("Haknovec", "Císarská rokle", "Kodská stena", "Radotánské údolí")) %>%
  pivot_wider(names_from = Name, values_from = pet) %>% 
  dplyr::select(-c(year, month)) %>%
  filter(complete.cases(.)) %>%
  cor(.)

# print("vapor pressure deficit")
# chelsa %>% 
#   dplyr::select(Name, month, year, vpd) %>%
#   filter(Name %in% c("Haknovec", "Císarská rokle", "Kodská stena", "Radotánské údolí")) %>%
#   pivot_wider(names_from = Name, values_from = vpd) %>% 
#   dplyr::select(-c(year, month)) %>%
#   filter(complete.cases(.)) %>%
#   cor(.)


```


# Climate loggers at the localities

```{r 2, echo= FALSE, warning=FALSE}

data_loggers_extract <- function(locality, data_loggers_folder = params$data_loggers_folder) {
  
  if(locality == "Ks") {  
    sheets <- as.list(paste0(locality, c(1,2)))
  } else {
    sheets <- as.list(paste0(locality, c(1:3)))
  }
  
  file <- paste0(locality, "_clima.xlsx")
  
  df <- lapply(sheets, function(x) readxl::read_xlsx(here::here(data_loggers_folder, file), sheet = x) %>% mutate(date = as.POSIXct(date, format = "%d.%m.%Y %H:%M:%S"))) %>% bind_rows(.id = "id") %>%
  mutate(id = as.factor(id)) %>%
  mutate(year = as.integer(format(date, format = "%Y")),
         month = as.integer(format(date, format = "%m")),
         day = as.integer(format(date, format = "%d"))) %>%
  group_by(id, year, month) %>%
  summarise(temp_below = mean(temp_below, na.rm = T),
            temp_surface = mean(temp_surface, na.rm = T),
            temp_above = mean(temp_abobe, na.rm = T),
            mean_moisture = mean(soil_moisture, na.rm = T),
            min_moisture = min(soil_moisture, na.rm = T)) %>%
  ungroup(year) %>%
  mutate(temp_below_scaled = scale(temp_below),
         temp_surface_scaled = scale(temp_surface),
         temp_above_scaled = scale(temp_above),
         mean_moisture_scaled = scale(mean_moisture),
         min_moisture_scaled = scale(min_moisture))
}

Cr <- data_loggers_extract("Cr")
Hk <- data_loggers_extract("Hk")
Ks <- data_loggers_extract("Ks")
Ru <- data_loggers_extract("Ru")

df <- bind_rows(list("Cr" = Cr, 
            "Hk" = Hk, 
            "Ks" = Ks, 
            "Ru" = Ru),
            .id = "locality")

ggplot(df) + 
  geom_line(aes(x = month, y = mean_moisture, colour = locality, linetype = id)) + 
  facet_wrap(vars(year))

ggplot(df) + 
  geom_line(aes(x = month, y = min_moisture_scaled, colour = locality, linetype = id)) + 
  facet_wrap(vars(year))

ggplot(df) + 
  geom_line(aes(x = month, y = temp_below_scaled, colour = locality, linetype = id)) + 
  facet_wrap(vars(year))

ggplot(df) + 
  geom_line(aes(x = month, y = temp_surface_scaled, colour = locality, linetype = id)) + 
  facet_wrap(vars(year))

ggplot(df) + 
  geom_line(aes(x = month, y = temp_above_scaled, colour = locality, linetype = id)) + 
  facet_wrap(vars(year))

```



